---
title: "Analysing the impact of deletion variants on susceptibility"
author: "Richard Goodman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysing the impact of deletion variants on susceptibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Exploring catB3 deletion variants and impact on chloramphenicol susceptibility in *Escherichia coli*

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  out.width = "100%"
)

# If the package is not installed, load the development version
if (!requireNamespace("AMRgen", quietly = TRUE)) {
  devtools::load_all()
}
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dpi = 72, dev = "png")
```

# 1. Introduction 

Infections caused by extended-spectrum beta-lactamase (ESBL)-producing Enterobacterales (ESBL-E) are a critical global health threat, often leaving clinicians with few treatment options beyond last-resort carbapenems. 

In Malawi, first-line treatment for sepsis shifted from chloramphenicol (CHL) to ceftriaxone (CRO) in 2004, this was followed by a notable re-emergence of CHL susceptibility as its clinical use declined ([Musicha et al., 2017](https://doi.org/10.1016/S1473-3099(17)30394-8)). 

A recent study ([Graf et al., 2024](https://doi.org/10.1038/s41467-024-53391-2)) revealed that this "resensitisation" is frequently driven by the stable degradation of resistance genes rather than their total loss from the population. Specifically, insertion sequences like IS*26* and IS*5* have been identified as key drivers; IS*26* causes truncations in *catB3* (creating the non-functional variant *catB4*), while IS*5* can integrate into the promoter of *catA1*, effectively silencing its transcription. 

Here we use a matched phenotype/genotype dataset used in the ([Graf et al., 2024](https://doi.org/10.1038/s41467-024-53391-2)) study and publicly available datasets from NCBI to investigate these genotype-phenotype mismatches with the AMRgen package. 

## 1.1 Sourcing data from the DASSIM study 

One of the datasets used to highlight genotype-phenotype mismatches in the [Graf et al., 2024](https://doi.org/10.1038/s41467-024-53391-2) paper was the DASSIM dataset ([Lewis et al. 2022](https://doi.org/10.1038/s41564-022-01216-7)).

The DASSIM study was an observational study of patients with sepsis admitted to Queen Elizabeth Central Hospital, Blantyre, Malawi. The aim was to understand the drivers of acquisition and long term carriage of ESBL-E in sepsis survivors. The DASSIM dataset contains faecal samples from community patients, inpatients and sepsis patients.

All genomic data was short-read sequenced on Illumina platforms at the Wellcome Sanger Institute. 

Antimicrobial sensitivity testing (AST) was carried out on a subset of isolates using the disc-diffusion method using British Society for Antimicrobial Chemotherapy (BSAC) guidelines (https://bsac.org.uk/). AST was carried out for meropenem, amikacin, chloramphenicol, ciprofloxacin, co-trimoxazole and gentamicin.

However this dataset only contains the interpreted phenotype data (S/I/R) which is what we work with in this example. 

## 1.1a DASSIM Genotype data 

We downloaded genome data from the following papers:

> [Colonization dynamics of extended-spectrum beta-lactamase-producing Enterobacterales in the gut of Malawian adults](https://doi.org/10.1038/s41564-022-01216-7)
Nat. Microbiol. 7, 1593–1604 (2022).
Joseph M Lewis, Madalitso Mphasa, Rachel Banda, Matthew Beale, Eva Heinz, Jane Mallewa, Christopher Jewell, Nicholas R Thomson, Nicholas A Feasey

> [Genomic analysis of extended-spectrum beta-lactamase (ESBL) producing Escherichia coli colonising adults in Blantyre, Malawi reveals previously undescribed diversity](https://doi.org/10.1099/mgen.0.001035)
Microb. Genom. 9, mgen001035 (2023).
Joseph M Lewis, Madalitso Mphasa, Rachel Banda, Matthew Beale, Jane Mallewa, Catherine Anscombe, Allan Zuza, Adam P Roberts, Eva Heinz, Nicholas Thomson, Nicholas A Feasey

> [Genomic and antigenic diversity of colonising Klebsiella pneumoniae isolates mirrors that of invasive isolates in Blantyre, Malawi](https://doi.org/10.1099/mgen.0.000778)
Microb. Genom. 8, 000778 (2022).
Joseph M Lewis, Madalitso Mphasa, Rachel Banda, Matthew Beale, Jane Mallewa, Eva Heinz, Nicholas Thomson, Nicholas A Feasey

The genomes are deposited in the European Nucleotide Archive (ENA) under the project IDs [PRJEB26677](https://www.ebi.ac.uk/ena/browser/view/PRJEB26677) and [PRJEB36486](https://www.ebi.ac.uk/ena/browser/view/PRJEB36486). 

The fastq files were downloaded, processed with cutadapt, filtered to >Q20 with FASTQC, and assembled with SPAdes v3.11.1 as described in [Graf et al., 2024](https://doi.org/10.1038/s41467-024-53391-2). 

Antimicrobial resistance genes (ARGs) were called with AMRFinderPlus v4.0.23 to produce the data frame [DASSIM_geno] which is included in the AMRgen package as a data object `DASSIM_geno`.

### 1.1b DASSIM Phenotype data 

We downloaded the AST data from the [blantyreESBL Github](https://joelewis101.github.io/blantyreESBL/) from Dr. Joe Lewis at <https://github.com/joelewis101/blantyreESBL/raw/refs/heads/main/data/btESBL_AST.rda>.

This is included in the AMRgen package as a data object: `btESBL_AST`.

Additional metadata relating to the isolates can be found in [Supplementary Data 1](https://static-content.springer.com/esm/art%3A10.1038%2Fs41467-024-53391-2/MediaObjects/41467_2024_53391_MOESM4_ESM.xls) of the following paper:

> [Molecular mechanisms of re-emerging chloramphenicol susceptibility in extended-spectrum beta-lactamase-producing Enterobacterales](https://doi.org/10.1038/s41467-024-53391-2). 
Nat Commun 15, 9019 (2024).
Fabrice E Graf, Richard N Goodman, Sarah Gallichan, Sally Forrest, Esther Picton-Barlow, Alice J Fraser, Minh-Duy Phan, Madalitso Mphasa, Alasdair T M Hubbard, Patrick Musicha, Mark A Schembri, Adam P Roberts, Thomas Edwards, Joseph M Lewis, Nicholas A Feasey.

This data is included in the AMRgen package as data object: `DASSIM_pheno_raw`.

# 2. Analysis 


## 2.1 Setting up R 

First we set up R and load our libraries. 

```{r, echo = T, error=FALSE, warning=FALSE, message=FALSE, results = 'hide'}
library(AMRgen)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## 2.2 Format the phenotype data

The phenotype data in `btESBL_AST` needs to be reformated to long format, and sequence identifiers imported from `DASSIM_pheno_raw` so we can match the phenotype data to the genotypes.

```{r, error=FALSE, warning=FALSE, message=FALSE}
# Convert the S/I/R phenotype data to long format for easy use with AMRgen functions
DASSIM_pheno <- btESBL_AST %>%
  pivot_longer(
    names_to = "drug",
    values_to = "pheno",
    cols = "amikacin":"meropenem"
  ) %>%
  mutate( # Standardise the terms to S, I, and R
    pheno = case_when(
      tolower(pheno) %in% c("sensitive", "susceptible", "s") ~ "S",
      tolower(pheno) %in% c("intermediate", "i") ~ "I",
      tolower(pheno) %in% c("resistant", "r") ~ "R",
      TRUE ~ NA_character_ # Any other unrecognised values become NA
    )
  ) %>%
  mutate(pheno = AMR::as.sir(pheno))

# add the sequence identifier from DASSIM_pheno_raw so we can match to genotype data
DASSIM_pheno <- DASSIM_pheno %>%
  left_join(DASSIM_pheno_raw %>% select(Strain_ID, seq, ST), join_by("supplier_name" == "Strain_ID")) %>%
  rename(id = seq) %>%
  relocate(id) %>%
  mutate(mic = NA)

head(DASSIM_pheno)
```

## 2.3 Check the genotype data

``` {r}
head(DASSIM_geno)
```


## 2.4 PPV Analysis of DASSIM dataset

The function `ppv()` predicts positive predictive value of genetic markers (i.e. genes/mutations) for resistance among strains that carry these markers. 

We will look at the markers associated with the antibiotics used in the AST assays for the DASSIM study. 

- chloramphenicol
- amikacin
- gentamicin
- cotrimoxazole
- meropenem

For each of these first we create a binary matrix using `get_binary_matrix()` which takes our genotype (AMRfinderplus etc.) and phenotype (AST profile) datasets. 

We can then plot the PPV graphs using `ppv()` 

```{r, fig.width=10, fig.height=5, error=FALSE, warning=FALSE, message=FALSE}
# Get binary matrix
DASSIM_CHL_bin_mat <- get_binary_matrix(DASSIM_geno, DASSIM_pheno, antibiotic = "chloramphenicol", sir_col = "pheno")
# Plot ppv
DASSIM_CHL_PPV <- ppv(DASSIM_CHL_bin_mat, antibiotic = "Chloramphenicol", sir_col = "pheno", upset_grid = FALSE)
```

We can see that *oqx* genes on their own don't impact resistance. As they are common, they obscure the impact of other genes, so we can remove them using `select()` to focus on the genes of interest for chloramphenicol.

```{r, fig.width=10, fig.height=5, error=FALSE, warning=FALSE, message=FALSE}
DASSIM_CHL_nooqx_bin_mat <- DASSIM_CHL_bin_mat %>% select(!contains("oqx"))
DASSIM_CHL_nooqx_PPV <- ppv(DASSIM_CHL_nooqx_bin_mat, antibiotic = "Chloramphenicol", sir_col = "pheno", upset_grid = FALSE)
```

This result clearly shows how the detection of *catB3* is not a good predictor of resistance, whereas detection of *catA2* is. 


Now let's check how well aminoglycoside markers predict resistance to amikacin and gentamicin.
```{r, fig.width=10, fig.height=5, error=FALSE, warning=FALSE, message=FALSE}
DASSIM_AMK_bin_mat <- get_binary_matrix(DASSIM_geno, DASSIM_pheno, antibiotic = "amikacin", sir_col = "pheno")
DASSIM_AMK_PPV <- ppv(DASSIM_AMK_bin_mat, antibiotic = "amikacin", sir_col = "pheno", upset_grid = TRUE)

DASSIM_GEN_bin_mat <- get_binary_matrix(DASSIM_geno, DASSIM_pheno, antibiotic = "gentamicin", sir_col = "pheno")
DASSIM_GEN_PPV <- ppv(DASSIM_GEN_bin_mat, antibiotic = "gentamicin", sir_col = "pheno", upset_grid = TRUE)
```

Here we see many of the aminoglycoside-associated resistance genes detected in these genomes are predictive of resistance to gentamicin, but none are associated with amikacin resistance. This is because amikacin is a semi-synthetic drug with an addition of a specific side chain, called the L-hydroxyaminobutyryl amide (HABA) group. This HABA side chain blocks the Aminoglycoside-Modifying Enzymes (AMEs) from reaching the sites on the molecule where they would normally attach their deactivating tags.


Now let's check how well markers associated with trimethoprims or sulfonamides predict resistance to co-trimoxazole.
```{r, fig.width=10, fig.height=5, error=FALSE, warning=FALSE, message=FALSE}
DASSIM_SXT_bin_mat <- get_binary_matrix(DASSIM_geno, DASSIM_pheno, antibiotic = "cotrimoxazole", sir_col = "pheno")
DASSIM_SXT_PPV <- ppv(DASSIM_SXT_bin_mat, antibiotic = "cotrimoxazole", sir_col = "pheno", upset_grid = TRUE)
```

Co-trimoxazole is a combination drug made of sulfamethoxazole and trimethoprim. It is prescribed prophylactically in Malawi for HIV-positive individuals ([Everett et al. 2011](https://doi.org/10.1371/journal.pone.0017765)). Since it is a combination drug it requires both Sulfamethoxazole resistance genes (e.g., *sul*) and Trimethoprim resistance genes (e.g., *dfrA*).

Meropenem is a last resort carbapenem antibiotic. ppv() by default returns all markers associated with beta-lactam resistance, for comparison with meropenem phenotypes. However while many beta-lactamases were detected, none are known carbapenemases (e.g., *bla~NDM~*, *bla~KPC~*, *bla~VIM~*, *bla~IMP~*, *bla~OXA-48-like~*). Consistent with this only a single isolate, carrying multiple beta-lactamases, was phenotyped as resistant to meropenem.
```{r, fig.width=13, fig.height=5, error=FALSE, warning=FALSE, message=FALSE}
DASSIM_MEM_bin_mat <- get_binary_matrix(DASSIM_geno, DASSIM_pheno, antibiotic = "meropenem", sir_col = "pheno")
DASSIM_MEM_PPV <- ppv(DASSIM_MEM_bin_mat, antibiotic = "meropenem", sir_col = "pheno", upset_grid = TRUE)
```

This analysis highlights how AMRgen can be used to explore genotype/phenotype associations for specific genetic markers related to a variety of antibiotics. 

Next we'll explore the chloramphenicol susceptibility associated with the *catB3* gene with more functions of the AMRgen package. For this we need raw AST data (e.g., MIC or disc diffusion), so we will go to NCBI to download public datasets. 

# 3. Analysis of publicly available phenotype and genotype data for chloramphenicol

## 3.1 Importing public data from NCBI

The National Center for Biotechnology Information (NCBI) provides tools for analysing antimicrobial resistance (AMR).

**Phenotypes: NCBI AST**

The Antibiotic Susceptibility Test (AST) Browser serves as a centralised resource for viewing and filtering phenotypic susceptibility data, allowing researchers to correlate specific bacterial isolates with their phenotypic resistance profiles.

AST data can be retrieved directly from NCBI using the AMRgen functions `download_ncbi_ast()` (slow but does not require authorisation) or `query_ncbi_bq_ast()` (very fast, requires a Google Cloud account), or via the NCBI AST Browser. For more details see the [DownloadGenoPhenoData] vignette.

``` {r download_ncbi_ast_notrun, eval=F}
# Download E. coli AST data from NCBI, filtering for chloramphenicol, and re-interpret with CLSI breakpoints
ecoli_ast_ncbi_via_biosample <- download_ncbi_ast(
  species = "E. coli",
  antibiotic = "chloramphenicol",
  reformat = TRUE,
  interpret_clsi = TRUE
)
```

```{r install_bigrquery, eval=F}
# Download E. coli AST data from NCBI via Google Cloud, filtering for chloramphenicol, and re-interpret with CLSI breakpoints

install.packages("bigrquery")
library(bigrquery)
bigrquery::bq_auth()

# replace xxx with your project id
ecoli_ast_ncbi_via_cloud <- query_ncbi_bq_ast(
  taxgroup = "E.coli and Shigella",
  antibiotic = "chloramphenicol",
  project_id = "xxx"
)

ecoli_ast_ncbi_via_cloud_interpreted <- import_ncbi_ast(ecoli_ast_ncbi_via_cloud,
  interpret_clsi = TRUE
)
```

Navigate to the [*NCBI  Antibiotic Susceptibility Test (AST) Browser*](https://www.ncbi.nlm.nih.gov/pathogens/ast) in a web browser and search for 

> `chloramphenicol AND Escherichia`

> https://www.ncbi.nlm.nih.gov/pathogens/ast#chloramphenicol%20AND%20Escherichia

Save this as a `tsv` file: `CHL_Ecoli_asts.tsv`, and import it using the AMRgen function `import_pheno`. 

```{r, error=FALSE, warning=FALSE, message=FALSE, eval=FALSE}
# import phenotype data
NCBI_Ecoli_AST_chl <- import_pheno("data-raw/CHL_Ecoli_asts.tsv", format = "ncbi")
```

A copy of this imported data (downloaded March 2026) is included in AMRgen as data frame `NCBI_Ecoli_AST_chl`.

```{r NCBI_Ecoli_AST_chl}
head(NCBI_Ecoli_AST_chl)
```

**Genotypes: MicroBIGG-E**

The Microbial Browser for Identification of Genetic and Genomic Elements (MicroBIGG-E) is a specialised portal within the NCBI Pathogen Detection system that enables users to query a database of over 46,000 isolates to identify specific AMR genes and point mutations. 

Navigate to the [*NCBI Pathogen Detection Microbial Browser for Identification of Genetic and Genomic Elements (MicroBIGG-E)*](https://www.ncbi.nlm.nih.gov/pathogens/microbigge) in a web browser and search for:

> `chloramphenicol AND Escherichia`

> https://www.ncbi.nlm.nih.gov/pathogens/microbigge/#chloramphenicol%20AND%20Escherichia

Save this as a `tsv` file: `CHL-R_Ecoli_microbigge.tsv`, and import it using the AMRgen function `import_gheno`.  

```{r, error=FALSE, warning=FALSE, message=FALSE, eval=FALSE}
# import phenotype data
MICROBIGGE_Ecoli_CHLR <- import_geno("data-raw/CHL-R_Ecoli_microbigge.tsv", format = "amrfp", sample_col = "BioSample")
```

A copy of this imported data (downloaded March 2026) is included in AMRgen as data frame `MICROBIGGE_Ecoli_CHLR`.

```{r MICROBIGGE_Ecoli_CHLR}
head(MICROBIGGE_Ecoli_CHLR)
```


## 3.2 Filter data to samples with chloramphenicol phenotype data, and chloramphenicol genotypic markers detected

``` {r filter}
# filter AST data, re-interpret using CLSI breakpoints
AST_pheno <- NCBI_Ecoli_AST_chl %>%
  filter(id %in% MICROBIGGE_Ecoli_CHLR$id) %>%
  interpret_ast(interpret_clsi = TRUE)
MB_CHLR_geno <- MICROBIGGE_Ecoli_CHLR %>% filter(id %in% NCBI_Ecoli_AST_chl$id)

# check how many samples we have
length(unique(AST_pheno$id))

# check the genes
MB_CHLR_geno %>% count(gene)

# filter to find samples with catB3
MB_CATB3_geno <- MB_CHLR_geno %>% filter(gene == "catB3")
MB_nonCATB3_geno <- MB_CHLR_geno %>% filter(gene != "catB3")

# filter AST data to samples with catB3 and no other chloramphenicol markers
AST_CATB3_pheno <- AST_pheno %>%
  filter(id %in% MB_CATB3_geno$id) %>%
  filter(!(id %in% MB_nonCATB3_geno$id))

# check how many samples we have with catB3
length(unique(AST_CATB3_pheno$id))
```

## 3.3 Exploring chloramphenicol phenotype distributions with `assay_by_var()`

Now we can explore phenotype distributions based on MIC data using the `assay_by_var()` function. 

We visualise the MIC distribution with `assay_by_var()` on all chloramphenicol AST data. 

```{r, error=FALSE, warning=FALSE, message=FALSE}
AST_pheno <- AST_pheno %>%
  mutate(across(all_of("mic"), ~ AMR::as.mic(.x, round_to_next_log2 = TRUE)))

assay_by_var(
  pheno_table = AST_pheno,
  antibiotic = "chloramphenicol",
  measure = "mic",
  colour_by = "pheno_clsi",
  species = "Escherichia coli"
)
```

Next we can visualise the MIC distribution with `assay_by_var()` on the AST data of isolates containing the *catB3* gene only

```{r, error=FALSE, warning=FALSE, message=FALSE}
# CATB3 specific

assay_by_var(
  pheno_table = AST_CATB3_pheno,
  antibiotic = "CHL",
  measure = "mic",
  colour_by = "pheno_clsi",
  species = "Escherichia coli"
) +
  labs(title = "Chloramphenicol MIC distribution for isolates with catB3 only")
```

The distribution shifts to the right and towards sensitive for isolates with the *catB3* gene only, when compared to those with any chloramphenicol-associated gene. 

We can then split the plot based on how truncated the *catB3* gene is as a percentage of coverage with 100% aligning across the entire length of the gene and <100% showing a truncation or deletion (see [Graf et al., 2024](https://doi.org/10.1038/s41467-024-53391-2) for more about the IS*26* mediated truncation of *catB3*)

```{r, fig.width=8, fig.height=10, error=FALSE, warning=FALSE, message=FALSE}
# add genotype data to the phenotype table for isolates with catB3 alone
AST_CATB3_pheno_2 <- MB_CATB3_geno %>%
  select(id, gene, `% Coverage of reference`) %>%
  distinct(id, .keep_all = TRUE) %>%
  right_join(AST_CATB3_pheno, by = "id")

# check coverage, all values are either 100% or 66.7-70%
AST_CATB3_pheno_2 %>% count(`% Coverage of reference`)

# define a grouping variable 'truncation' indicating samples with full coverage vs <=70%
AST_CATB3_pheno_3 <- AST_CATB3_pheno_2 %>%
  mutate(truncation = ifelse(`% Coverage of reference` > 70, "100% catB3 coverage", "<= 70 % catB3 coverage"))

# plot the MIC distribution for these 2 groups
MIC_dist_by_cov <- assay_by_var(
  pheno_table = AST_CATB3_pheno_3,
  antibiotic = "CHL",
  measure = "mic",
  colour_by = "pheno_clsi",
  species = "Escherichia coli",
  facet_var = "truncation"
) +
  labs(
    title = " ",
    subtitle = " ",
    x = "MIC measurement",
    fill = "phenotype (CLSI)"
  )

MIC_dist_by_cov

# check counts and median MIC per group
AST_CATB3_pheno_3 %>%
  group_by(truncation) %>%
  summarise(median = median(mic, na.rm = T), n = n(), R = sum(pheno_clsi == "S"))
```

As we can see, isolates with truncated *catB3* genes (n=110) have median MIC of 4 mg/L, and most (n=100) were classed as susceptible (MIC <16 mg/L). In contrast, those with full-length *catB3* genes (n=6) had higher values, and only 3 were classed as susceptible.

## 3.4 Analysing genotype and phenotype data with `ppv()`

Now let's look at the geno-pheno associations across all the isolates with matched data. We first build a binary matrix using the genotype and phenotype tables as input with `get_binary_matrix()`

Then we can view it as an upset grid, with a upset plot, SIR stacked barplot and positive predictive value (ppv) using the `ppv()` function. 

```{r, fig.width=10, fig.height=5, error=FALSE, warning=FALSE, message=FALSE}
MB_CHLR_geno <- MB_CHLR_geno %>%
  mutate(marker = ifelse(marker == "catB3" & `% Coverage of reference` < 100, "catB3 (truncated)", marker))

CHL_bin_mat <- get_binary_matrix(MB_CHLR_geno,
  AST_pheno,
  antibiotic = "CHL",
  drug_class_list = c("Phenicols"),
  sir_col = "pheno_clsi",
  keep_assay_values = TRUE
)

CHL_PPV <- ppv(CHL_bin_mat,
  antibiotic = "Chloramphenicol",
  drug_class_list = c("Phenicols"),
  sir_col = "pheno_clsi",
  upset_grid = TRUE,
  assay = "mic",
  plot_assay = TRUE,
  order = "value"
)
```
## 3.5 Logistic regression

We can plot a coefficient plot using the `amr_logistic()` function to show the statistical relationship between chloramphenicol resistance genes and phenotypic resistance to chloramphenicol. 

We just need the binary matrix as input (from `get_binary_matrix()`) and we can define the antibiotic, the column containing the ecoff and a threshold for the number of samples a marker is present in. 

```{r, error=FALSE, warning=FALSE, message=FALSE}
CHL_logist <- amr_logistic(
  binary_matrix = CHL_bin_mat,
  antibiotic = "chloramphenicol",
  ecoff_col = "ecoff",
  maf = 10,
  single_plot = TRUE
)

# model coefficients
CHL_logist$modelR
```
This plot shows that both *catA1* and *cmlA1* have a strong positive association with phenotypic resistance whereas truncated *catB3* has a strong negative association with resistance (i.e. a stronger association with susceptibility to chloramphenicol). 

Experimental evolution analysis by [Graf et al., 2024](https://doi.org/10.1038/s41467-024-53391-2) suggested that these *catB3* silencing events are highly stable under antibiotic pressure, reinforcing the potential for chloramphenicol to be reintroduced as a targeted reserve drug for ESBL-E infections in low-resource settings.
