---
title: "Download NCBI and EBI Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Download NCBI and EBI Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  out.width = "100%"
)
```

# Introduction
This vignette demonstrates how to download antibiotic susceptibility testing (AST) data from NCBI and EBI, and to re-interpret it using different clinical breakpoints. 

Start by loading the `AMRgen` package:

```{r setup}
library(AMRgen)
library(dplyr)
```

## 1. Download data from NCBI
The `download_ncbi_ast()` function lets you download antibiogram data from NCBI. You must specify a species, and can optionally limit the download to one or more specific drugs. The function can also re-format the data into an AMRgen phenotype table, and re-interpret phenotypes against clinical breakpoints from EUCAST or CLSI.

``` {r download_ncbi_data_realnotrun, eval=F}
# Download Staphylococcus aureus AST data from NCBI, filtering for ampicillin, and re-interpreting with EUCAST breakpoints
staph_ast_ncbi <- download_ncbi_ast(
  species = "Staphylococcus aureus",
  antibiotic = c("amikacin", "DOX"), # antibiotics can be listed in short or long form
  reformat = TRUE,
  interpret_eucast = TRUE
) # reformat must be true to use interpret_* argument
```

``` {r download_ncbi_data_dummy}
# check how many samples retrieved
nrow(staph_ast_ncbi)

# check the output
head(staph_ast_ncbi)
```

``` {r download_ncbi_dataraw_realnotrun, eval=F}
# This is the same as downloading the data then re-interpreting it separately:
staph_ast_ncbi_raw <- download_ncbi_ast(
  species = "Staphylococcus aureus",
  antibiotic = c("amikacin", "DOX"),
  reformat = FALSE,
  interpret_eucast = FALSE
)
```

``` {r download_ncbi_dataraw_dummy}
head(staph_ast_ncbi_raw)

# Then reformat and re-interpret using EUCAST and CLSI breakpoints, and ECOFFs using the import_ncbi_biosample() function
staph_ast_ncbi2 <- import_ncbi_biosample(
  input = staph_ast_ncbi_raw,
  interpret_clsi = TRUE,
  interpret_eucast = TRUE,
  interpret_ecoff = TRUE
)

head(staph_ast_ncbi2)

# Note that when there is a mix of MIC and disk data, a separate _disk and _mic interpretation column, as well as an overall phenotype column, is produced for each interpretation.
```
This produces a long format data frame, with one row per sample and drug combination. This is compatible with downstream functions in the AMRgen package.

Consider altering the `max_records`, `batch_size`  or `sleep_time` options if you want to download a lot of data or run into NCBI server issues. 

### Visualise the downloaded phenotype data to check the distribution of the AST data
``` {r assay_by_var_staph_doxy}
# Select one antibiotic at a time
# Specify species and guideline, to annotate with EUCAST breakpoints

# Doxycycline
staph_dox_mic_plot <- assay_by_var(
  pheno_table = staph_ast_ncbi,
  antibiotic = c("DOX"),
  measure = "mic",
  colour_by = "pheno_eucast",
  species = "Staphylococcus aureus",
  guideline = "EUCAST 2024"
)

staph_dox_mic_plot
```

``` {r assay_by_var_staph_ami}
# Amikacin
staph_ami_mic_plot <- assay_by_var(
  pheno_table = staph_ast_ncbi,
  antibiotic = c("amikacin"),
  measure = "mic",
  colour_by = "pheno_eucast",
  species = "Staphylococcus aureus",
  guideline = "EUCAST 2024"
)

staph_ami_mic_plot
```
For more guidance on how to visulaise phenotypic data and combine it with genotypic data, have a look at the other `AMRgen` [vignettes](https://amrverse.github.io/AMRgen/articles/AnalysingGenoPhenoData.html).


## 2. Download data from EBI
The `download_ebi` function lets you retrieve phenotype or genotype data (by setting `data ="genotype'`). You can optionally filter the downloaded file to a specified genus or species, and a specific antibiotic (for phenotype data) or NCBI class/subclass (for genotype data; check the [NCBI AMR Class-Subclass Reference](https://github.com/ncbi/amr/wiki/class-subclass) for valid terms). The function can also reformat and re-interpret the retrieved results with the breakpoint standards of your choosing.

``` {r download_ebi_ast_realnotrun, eval=F}
# Download EBI phenotype data for all Staphylococcus, using the same example drugs as above. Reformat and re-interpret using EUCAST breakpoints
staph_ast_ebi <- download_ebi(
  genus = "Staphylococcus",
  antibiotic = c("amikacin", "DOX"),
  reformat = TRUE,
  interpret_eucast = TRUE,
  interpret_clsi = TRUE,
  interpret_ecoff = TRUE
) # chose which guideline to use for re-interpretation. reformat must be TRUE for re-interpretation
```

``` {r download_ebi_ast_dummy}
# check output
nrow(staph_ast_ebi)

length(unique(staph_ast_ebi$id))

head(staph_ast_ebi)
```

``` {r download_ebi_geno_realnotrun, eval=F}
# Download all available genotype data for Staphylococcus, and re-format the data into an AMRgen genotype table. Note that not all samples with phenotype data have genotype data.
staph_geno_ebi <- download_ebi(
  data = "genotype",
  genus = "Staphylococcus",
  reformat = T
)
```

``` {r download_ebi_geno_dummy}
nrow(staph_geno_ebi)

length(unique(staph_geno_ebi$BioSample_ID))

head(staph_geno_ebi)
```

### Compare downloaded phenotypes and genotypes
The downloaded phenotype and genotype data for a specified antibiotic can then be extracted and combined using the `get_binary_matrix` function.
```{r get_binary_matrix}
# first filter both EBI pheno and geno dataframes for Staph aureus only
# filter pheno data
staph_ast_ebi_filtered <- staph_ast_ebi %>%
  filter(organism == "Staphylococcus aureus")

# filter geno data
staph_geno_ebi_filtered <- staph_geno_ebi %>%
  filter(species == "Staphylococcus aureus")

# Make binary geno-pheno matrix for doxycycline phenotype (re-interpreted with EUCAST), and genotypes associated with the associated drug class (Tetracyclines)
tet_bin <- get_binary_matrix(
  geno_table = staph_geno_ebi_filtered,
  pheno_table = staph_ast_ebi_filtered,
  antibiotic = "DOX",
  drug_class_list = "Tetracyclines", # matches drug_class in geno_table
  sir_col = "pheno_eucast", # phenotype column in pheno_table
  keep_assay_values = TRUE,
  keep_assay_values_from = "mic"
)

nrow(tet_bin)

head(tet_bin)
```

``` {r ppv, fig.height=8}
# plot positive predictive value for each marker/combination
tet_ppv <- ppv(tet_bin)

tet_ppv$plot
```
This produces a binary matrix with 1 row per BioSample, for all BioSamples that had both phenotype data for doxycycline in `staph_ast_ebi_filtered` AND any genotype data (associated with any marker, not just Tetracyclines) in `staph_geno_ebi_filtered`. This ensures that we include samples for which genotyping was performed but returned no hits associated with Tetracyclines, not just those samples that had tetracycline associated markers detected (note that such samples would be missing if we had only downloaded genotype data associated with Tetracyclines).

Columns indicate the input phenotype columns for doxycycline (renamed to standard fields `pheno`, `ecoff`, `mic`), and binary indicators (1=present, 0=absent) for doxycycline phenotypes (`R`, `NWT`) and each genotype marker associated with tetracyclines (in this case `tet(38)`, `tet(K)`, `tet(M)`,`tet(L)`).

For more examples on how to do join geno-pheno analyses and visualisations, so the other AMRgen [Vignettes](https://amrverse.github.io/AMRgen/articles/AnalysingGenoPhenoData.html).
