---
title: "Example with custom classification of genotype hits"
author: "Natacha Couto"
date: "2026-03-12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example with custom classification of genotype hits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Analysing clindamycin resistance in *Staphylococcus aureus*

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 15,
  fig.height = 15,
  fig.retina = 2,
  dpi = 300,
  out.width = "100%"
)
# If the package is not installed, load the development version
if (!requireNamespace("AMRgen", quietly = TRUE)) {
  devtools::load_all()
}
```

## Introduction

This is document summarises the analysis of clindamycin resistance in *S. aureus*. More specifically, we will look at the distribution of clindamycin resistance in *S. aureus* isolates from a dataset of clinical samples.

Start by loading the package:

```{r setup, warning=FALSE, message=FALSE}
library(AMRgen)
library(ggplot2)
library(dplyr)
```

## Data preparation

For this example, we have collated genotype-phenotype data from NCBI and EBI through the ESGEM-AMR Staphylococcus subgroup. Phenotypic data were collated into a single table (one row per isolate). The genotype data was generated using AMRFinderPlus. The pre-loaded objects `pheno_CLI_public` and `afp_CLI_public` serve as the **input** for `AMRgen`. First, we will create a new `variant.label` column that includes the marker gene name and its closest accession number, and also indicate whether the match to the accession is exact or not.

```{r extract subvariants, warning=FALSE, message=FALSE}
# Create the new marker.label column
afp_CLI_public <- afp_CLI_public %>%
  mutate(exact_match = if_else(
    `% Coverage of reference sequence` == 100 &
      `% Identity to reference sequence` == 100.00,
    "", "x_"
  )) %>%
  mutate(node_hit = paste0(node, "__", exact_match, `Accession of closest sequence`)) %>%
  mutate(variant.label = case_when(
    `variation type` == "Inactivating mutation detected" ~ paste0(node_hit, ":-"),
    !is.na(mutation) ~ paste0(node_hit, ":", mutation),
    TRUE ~ node_hit
  ))
```

## Genotype-phenotype analysis

Now let's create some upset plots, to see how genes, and their allelic variants, relate to clindamycin MICs.

```{r upset_gene, warning=FALSE, message=FALSE}
# Visualise with UpSet plot (markers)
cli_mic_upset <- amr_upset(
  geno_table = afp_CLI_public,
  pheno_table = pheno_CLI_public,
  marker_col = "marker.label",
  pheno_drug = "Clindamycin",
  min_set_size = 2,
  assay = "mic",
  order = "value",
  print_set_size = TRUE,
  plot_set_size = TRUE,
  print_category_counts = TRUE,
  bp_S = 0.25,
  bp_R = 0.5
)
```

```{r upset_variant, warning=FALSE, message=FALSE}
# Visualise with UpSet plot (markers)
# order markers alphabetically so we can more easily see where there are different variants of the same gene
cli_mic_upset_variant <- amr_upset(
  geno_table = afp_CLI_public,
  pheno_table = pheno_CLI_public,
  marker_col = "variant.label",
  pheno_drug = "Clindamycin",
  min_set_size = 2,
  assay = "mic",
  order = "value",
  marker_order = "alpha", # order markers alphabetically
  print_set_size = TRUE,
  plot_set_size = TRUE,
  print_category_counts = TRUE,
  bp_S = 0.25,
  bp_R = 0.5
)
```

## Solo PPV analysis

Next, we will calculate the solo positive predictive value (soloPPV) for each marker and variant. This will allow us to see which markers and variants are most predictive of clindamycin resistance.

```{r visualise upset plot, warning=FALSE, message=FALSE}
# soloPPV analysis for markers
# order markers alphabetically so we can more easily see where there are different variants of the same gene
cli_soloPPV <- solo_ppv(
  geno_table = afp_CLI_public,
  pheno_table = pheno_CLI_public,
  marker_col = "marker.label",
  pheno_drug = "Clindamycin",
  order_ppv = FALSE
)

# soloPPV analysis for variants
# order markers alphabetically so we can more easily see where there are different variants of the same gene
cli_soloPPV_variant <- solo_ppv(
  geno_table = afp_CLI_public,
  pheno_table = pheno_CLI_public,
  marker_col = "variant.label",
  pheno_drug = "Clindamycin",
  order_ppv = FALSE
)

cli_soloPPV_variant$solo_stats
```

We can see from this analysis that the variants of the markers have different associations with clindamycin resistance. For example, the isolates with exact matches to erm(C)\_WP001003263.1 or erm(C)\_WP001003264.1 are associated with low PPV for resistance (8.5%, n=60/705 and 17%, n=2/12, respectively). This highlights the importance of considering the specific variants of the markers when predicting clindamycin resistance. Note: this discrepancy may be caused by inducible resistance, which is not always captured by a standard AST test. This is because the resistance gene may not be expressed under the conditions of the AST test, but can be induced in the presence of certain antibiotics (e.g., erythromycin). This is a known phenomenon for clindamycin resistance in *S. aureus*, where the presence of an erm gene can lead to inducible resistance that may not be detected in a standard AST test.

## MIC distribution associated with allelic variants
We can use the `assay_by_var()` function to visualise the MIC distributions associated with different variants of the same gene. First we need to join the key genotype fields into the phenotype table, and then use `assay_by_var()` to plot MIC assay measures grouped by variant and faceted into one panel per gene, coloured by S/I/R phenotype.
``` {r}
cli_geno_pheno <- afp_CLI_public %>%
  filter(`variation type` == "Gene presence detected") %>%
  mutate(variant_hit = if_else(exact_match == "",
    `Accession of closest sequence`,
    paste0(`Accession of closest sequence`, "_x")
  )) %>%
  select(id, marker.label, variant_hit) %>%
  left_join(pheno_CLI_public)


cli_mic_byhit <- assay_by_var(cli_geno_pheno, group_by = "variant_hit", 
                              boxplot = T, colour_by="pheno_eucast")$plot +
  facet_wrap(~marker.label, scales = "free_y", ncol = 2) +
  coord_flip() +
  theme(legend.position = "none")

cli_mic_byhit + ggtitle(expression(paste(
  "Clindamycin MIC in ",
  italic("S. aureus"),
  ", by gene symbol and closest reference"
)))
```

## Association of specific variants with sequence types

Finally, we can also look at the association of specific marker variants with sequence types, if these data are available. This can be done by creating a binary matrix for the variants and then visualizing the associations with a bubble plot.

```{r visualise bubble plot, warning=FALSE, message=FALSE}
# Identify "High-Frequency" STs because we have many STs with only a few samples, which can make the plot cluttered and less informative. By filtering to include only STs with a certain number of samples (e.g., n >= 100), we can focus on the most prevalent STs in the dataset, which are likely to provide more meaningful insights into the associations between specific variants and sequence types.
high_freq_sts <- ST_data_CLI %>%
  count(ST) %>%
  filter(n >= 100) %>%
  pull(ST)

# Merge with ST data
cli_accession_st <- afp_CLI_public %>%
  inner_join(ST_data_CLI) %>%
  filter(ST %in% high_freq_sts)

# Calculate prevalence of variant in each ST
balloon_data <- cli_accession_st %>%
  group_by(ST, variant.label) %>%
  summarise(n_samples = n_distinct(id), .groups = "drop") %>%
  left_join(ST_data_CLI %>% count(ST, name = "total_st_n"), by = "ST") %>%
  mutate(percent_prevalence = (n_samples / total_st_n) * 100) %>%
  filter(percent_prevalence >= 10)

# Create the Balloon Plot showing only marker-ST combinations with at least 10% prevalence
ggplot(balloon_data, aes(x = ST, y = variant.label)) +
  geom_point(aes(size = n_samples, color = percent_prevalence)) +
  scale_color_viridis_c(option = "viridis", direction = -1) +
  scale_size_continuous(range = c(1, 10)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 16),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 14)
  ) +
  labs(
    title = "Clindamycin Markers by Major Sequence Types",
    subtitle = "Analysis of STs with n >= 100",
    x = "Sequence Type (ST)",
    y = "Marker variant",
    size = "Sample Count",
    color = "Prevalence (%)"
  )
```

We can see from this last figure, that while erm(C)\_WP0012364.1 is quite rare, erm(C)\_WP0012363.1 is prevalent in certain STs, such as ST22. This suggests that there may be a clonal association of this variant with this ST, indicating resistance in this clone might be inducible and therefore more difficult to identify using standard AST.
